Multi-level modeling with Openflexo/\FML

### 2022-03-16 6-modelpresentation.tex
typos, missing reference, rephrasing

The designed multilevel architecture shown in
Fig.~\ref{fig:MultilevelArchitecture} captures the two use cases described in
the Process Challenge.

Note that in all following figures, blue is used to represent \FML virtual
models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in
Fig.~\ref{fig:mm}), see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:BaseMetamodel},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture},
while brown is used for instances of \textsf{FlexoConcept}, see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9 Expliquer que le use case XSure ne nécessitait pas
% l'extension de metamodèle requise pour le usecase Acme Software Developement
% Process

@@

Our proposition relies on ontological instantiation as presented in
Fig.~\ref{fig:LinguisticAndOntologicInstantiation}, with a common root concept
\texttt{ModelingElement}. Two types of ontological instantiation are needed to
meet the requirements of the challenge. One provides that some instances
conform to only one type (\texttt{Process} and \texttt{Task} do respectively
conform to \texttt{ProcessType} and \texttt{TaskType}). While some entities
define their conformity to several types (\texttt{Actor} and \texttt{Artifact}
do respectively conform to several \texttt{ActorType} and
\texttt{ArtifactType}). These instantiations are respectively expressed using
the relations \texttt{type} and \texttt{types} between \texttt{Type},
\texttt{Instance} and \texttt{MultiInstance} concepts. In the diagram, the
realization of these relations are indicated by a derived relation,
\texttt{/type} or \texttt{/types}, as for example between \texttt{Process} and
\texttt{ProcessType} or between \texttt{Actor} and \texttt{ActorType}.

@@

We first present the process definition part of the base metamodel, located on
the left of Fig.~\ref{fig:BaseMetamodel}.

@@

\subsubsection{Process enactment}
\label{sec:ProcessEnactment}
We now present the process enactment part of base metamodel, located right of
Fig.~\ref{fig:BaseMetamodel}. All concepts defined in this subsection are
either specializations of the \texttt{Instance} concept (if they have exactly
one type) or the \texttt{MultiInstance} concept (when they have several types).

@@

The Fig.~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the
architecture. This figure is only partial at instance level and not all
instances are represented. It also contains an instance of enacted a
\textit{Software Engineering Process} at the bottom. The Acme metamodel
specializes some concepts of the Base metamodel, for example,
\texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType}
extends \texttt{ArtifactType}. These concepts are further specialized
\texttt{CodingTaskType} extends \texttt{SETaskType} and
\texttt{CodeArtifactType} extends \texttt{SEArtifactType}.  The Acme metamodel
also defines a specialized task type \texttt{Coding} extending \texttt{Task},
and provides the \texttt{Developer} concept extending \texttt{ActorType}. This
metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined
as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the
instances required to capture the challenge use case are defined in the final
Acme model (lower part of the figure). Instances are represented with rounded
boxes and linguistic instantiation is represented using dashed connectors.

>>>>

The designed multilevel architecture shown in
Fig.~\ref{fig:MultilevelArchitecture} captures the two use cases described in
the Process Challenge.
Note that in all following figures, blue is used to represent \FML virtual
models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in
Fig.~\ref{fig:mm}), see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:BaseMetamodel},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture},
while brown is used for instances of \textsf{FlexoConcept}, see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9 Expliquer que le use case XSure ne nécessitait pas
% l'extension de metamodèle requise pour le usecase Acme Software Developement
% Process

@@

Our proposition relies on ontological instantiation as presented in
Fig.~\ref{fig:LinguisticAndOntologicInstantiation}, with a common root concept
\texttt{ModelingElement}. Two types of ontological instantiations are needed to
meet the requirements of the challenge. One provides that some instances
conform to only one type (\texttt{Process} and \texttt{Task} do respectively
conform to \texttt{ProcessType} and \texttt{TaskType}). While some entities
define their conformity to several types (\texttt{Actor} and \texttt{Artifact}
do respectively conform to several \texttt{ActorType} and
\texttt{ArtifactType}). These instantiations are respectively expressed using
the relations \texttt{type} and \texttt{types} between \texttt{Type},
\texttt{Instance} and \texttt{MultiInstance} concepts. In the diagram, the
realization of these relations are indicated by a derived relation,
\texttt{/type} or \texttt{/types}, as for example between \texttt{Process} and
\texttt{ProcessType} or between \texttt{Actor} and \texttt{ActorType}.

@@

In this subsection, we present the process definition part of the base
metamodel, located on the left of Fig.~\ref{fig:BaseMetamodel}.

@@

\subsubsection{Process enactment}
\label{sec:ProcessEnactment}
In this subsection, we present the process enactment part of base metamodel,
located right of Fig.~\ref{fig:BaseMetamodel}. All concepts defined in this
subsection are either specializations of the \texttt{Instance} concept (if they
  have exactly one type) or the \texttt{MultiInstance} concept (when they have
  several types).

@@

The Fig.~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the
architecture. This figure is only partial at instance level and not all
instances are represented. It also contains an instance of enacted a
\textit{Software Engineering Process} at the bottom. The Acme metamodel
specializes some concepts of the Base metamodel, for example,
\texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType}
extends \texttt{ArtifactType}. These concepts are further specialized
\texttt{CodingTaskType} extends \texttt{SETaskType} and
\texttt{CodeArtifactType} extends \texttt{SEArtifactType}.  The Acme metamodel
also defines a specialized task type \texttt{Coding} extending \texttt{Task},
and provides the \texttt{Developer} concept extending \texttt{ActorType}. This
metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined
as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the
instances required to capture the challenge use case are defined in the final
Acme model (lower part of the Fig.~\ref{fig:AcmeFullArchitecture}). Instances
are represented with rounded boxes and linguistic instantiation is represented
using dashed connectors.
